Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden, England and Wales

### 2020-08-25 reports/article.Rmd
Stat figs. Fix #52.

We used Kaplan--Meier and a procedure with bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015] Age and BMI were normalized to have mean = 0 and standard deviation = 1. Co-morbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process. 1,000 bootstrap samples were drawn from the observed data set. We used 10-fold cross validation for every bootstrap sample with a range of potential penalty values ($\lambda$:s) in a logistic LASSO model. We  kept $\lambda$:s minimizing the mean cross-validated deviances and those were used to estimate model coefficients for each potential predictor. The magnitude (absolute values) of those estimates were used as a measure of variable importance. Piece-wise linear regression was used to detect a breaking point where a significant drop in variable importance was observed. Potential predictors with variable importance above this breaking point were kept as model candidates. The whole process was repeated 100 times. Covariates that were selected at least once were used in a main effects model of multivariable logistic regression without penalty, and without pre-normalization of numeric variables (main model). A reduced model with variables chosen at least 33 out of the 100 times were used as a simpler alternative. This model was also evaluated without cancer as a predictor, considering that medical indications for THA surgery may be different for patients with cancer or without cancer. Univariable models with the ASA class, Charlson or Elixhauser co-morbidity indices were used for benchmarking, as well as a model including only age and sex. Each model including age was fitted three times, once with age as a main effect and twice with restricted cubic splines, either by two or three knots. Odds ratios for the final model were estimated with 95 % confidence intervals and $p$-values below 0.05 were considered statistically significant.

>>>>

We used a statistical procedure with bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO; ~~figure \@ref(fig:br1) and \@ref(fig:br2)~~).[@Guo2015] Age and BMI were normalized to have mean = 0 and standard deviation = 1. Co-morbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process. 1,000 bootstrap samples were drawn from the observed data set. We used 10-fold cross validation for every bootstrap sample with a range of potential penalty values ($\lambda$:s) in a logistic LASSO model. We  kept $\lambda$:s minimizing the mean cross-validated deviances and those were used to estimate model coefficients for each potential predictor. The magnitude (absolute values) of those estimates were used as a measure of variable importance. Piece-wise linear regression was used to detect a breaking point where a significant drop in variable importance was observed. Potential predictors with variable importance above this breaking point were kept as model candidates. The whole process was repeated 100 times. Covariates that were selected at least once were used in a main effects model of multivariable logistic regression without penalty, and without pre-normalization of numeric variables (main model). A reduced model with variables chosen at least 33 out of the 100 times were used as a simpler alternative. This model was also evaluated without cancer as a predictor, considering that medical indications for THA surgery may be different for patients with cancer or without cancer. Univariable models with the ASA class, Charlson or Elixhauser co-morbidity indices were used for benchmarking, as well as a model including only age and sex. Each model including age was fitted three times, once with age as a main effect and twice with restricted cubic splines, either by two or three knots. Odds ratios for the final model were estimated with 95 % confidence intervals and $p$-values below 0.05 were considered statistically significant.

### 2020-08-25 reports/figs.Rmd
Stat figs. Fix #52.

>>>>

(ref:br1) ~~Outer steps of the bootstrap ranking procedure. All included patients (A) were re-sampled with replacements 100 times (B). For each bootstrap sample, the inner process (C; as depicted in figure \@ref(fig:br2)) was applied. The final output was a table with all potential predictors ranked by the number of times they got selected (D). Potential predictors selected at least once were included in the main model, and each predictor selected at least 33 out of the 100 times, were included in a reduced model as well.~~

```{r br1, fig.cap = "(ref:br1)", out.width = "\\columnwidth"}
# knitr::include_graphics("../graphs/br1.pdf")
knitr::include_graphics("../graphs/images.png")
```

(ref:br2) ~~Inner steps of the bootstrap ranking procedure. For each bootstrap sample (1), 100 new bootstrap samples were created (2). Logistic LASSO-regression was applied within each sample (3). Absolute values of the estimated coefficient values (whereof some shrunk to 0 by the LASSO), were averaged as a measure of variable importance (4). Variables with their estimated variable importance above an estimated break-point from linear piece-wise regression (5) were kept as potential predictors (6).~~

```{r br2, fig.cap = "(ref:br2)", out.width = "\\columnwidth"}
# knitr::include_graphics("../graphs/br2.pdf")
knitr::include_graphics("../graphs/images.png")
```
